Freezing of a Weeks-Chandler-Anderson Fluid using classical Density Functional Theory

### 2020-11-10 papers/thesis-kirstie/thesis.tex
edits

%A number density profile is a time-invariant average 
%of the spacial distribution of the particles that make up the fluid, and
%There will be one, and only one 
%that minimizes the Helmholtz free energy.
%The minimum free energy, 
%is the value of the Helmholtz free energy that
%the system will have when it is in equilibrium at that temperature, 
%and its corresponding number density profile 
%are the equilibrium values the system will have at that temperature.
%is the equilibrium number density profile. 
%
In this section 
%To form the functional for the Helmholtz free energy, expressions for 
a functional for the Helmholtz free energy was formed by adding 
the ideal and the excess portions of the free 
energy.
\begin{equation}{F[n(\vec{r})]=F_{ideal}[n(\vec{r})] + F_{excess}[n(\vec{r})]}\end{equation}
The functional for the ideal portion of the Helmholtz free energy is given by 
Equation~\ref{eq:Fideal-n(r)}, and a functional for the excess portion of the Helmholtz 
free energy is given in Equation~\ref{eq:Fexcess-simplified}. The specific 
form for the excess Helmholtz free energy depends on the particular 
fluid being modeled as defined by the  
interatomic potential energy V(r) between the spheres, which can make
forming an expression difficult. 
Once an expression for the Helmholtz free energy as a functional of the
number density profile $n(\vec r)$ is formed then, according to classical 
Density Functional Theory, the equilibrium value of the 
Helmholtz energy, and its cooresonding equilibrium density profile can be
found by varying the number density profile
%can be varied 
through all possible configurations at some fixed temperature
until the free energy is minimized.

\section{Soft Fundamental Measure Theory (SFMT)}
In order to use classical Density Functional Theory, a functional for 
the excess Helmholtz free
energy for the particular fluid being modeled must be formed. But forming 
an expression based on Equation~\ref{eq:Fexcess-simplified} can be challenging. 
For the Hard Sphere fluid, there is another way to form an expression for 
the excess Helmholtz free energy in terms of a homogeneous number density 
and, that is to use Scaled Particle Theory (SPT)~\cite{ReissSPT}.

Scaled Particle Theory seeks a general expression for the excess chemical 
potential of a homogeneous Hard Sphere fluid which can then be integrated 
to find the excess Helmholtz free energy.
%\begin{equation}F_{ex}=\int{\mu_{ex}dN}\end{equation}
%\begin{equation}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}\end{equation}
\begin{equation}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}{~~~~~}\rightarrow{~~~~~}F_{ex}=\int{\mu_{ex}dN}\end{equation}
It is an extension of the work done by Widom who showed that the excess 
chemical potential of a homogenesous fluid corredsponds to the reversible 
work of inserting a sphere of radius R into a fluid consisting of hard 
spheres of radius R. Widom's result was obtained by considering the difference 
in the excess Helmholtz free energy when one atom is added to an otherwise 
closed system. He showed that the excess chemcial potential is related to 
the ensemble average of the Boltzman factor for the potential energy 
difference $V_{diff}$ between a system with N atoms and one with N+1 atoms. 
The resulting equation is known as Widom's Insertion Formula given by
%\begin{displaymath}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}{~~~~~~}\end{displaymath}
\begin{equation}\mu_{ex}=\frac{F_{ex}(N+1)-F_{ex}(N)}{\Delta{N}}\bigg|_{T,V}\end{equation}
\begin{equation}\label{widoms-insertion-formula}{~}=-k_BT\ln\left(<e^{-\beta{V_{diff}}}>\right)\end{equation}
\color{red} \begin{equation}=-k_BT\ln(1-n\frac{4\pi}{3}R_{cavity}^3)\end{equation}
where the radius of the cavity $R_{cavity}$ is equal to the radius of a 
hard sphere, R, and the cavity is either filled or not. \color{black}
%For cavities that are much larger than one of the hard spheres, 
Forming a more general result for different sizes of cavities in the fluid, 
Scaled Partical Theory arrives at an expression for the excess Helmholtz 
free energy of a homogeneous fluid~\cite{Hansen} in terms of the packing fraction $\eta$ 
\begin{equation}\label{Fexcess-SPT}{\frac{\beta{F_{excess}}}{N}=-ln(1-\eta)+\frac{3\eta}{1-\eta}+\frac{3{\eta}^2}{2(1-\eta)^2}}\end{equation} 
where
\begin{equation}{\eta = \frac{\mbox{Volume occupied by the spheres}}{\mbox{Total Volume}}}\end{equation}
\begin{equation}{\eta = \frac{N\frac{4}{3}\pi{R^3}}{V}}\end{equation}

@@

Incorporating into Equation~\ref{FexfromSPT} the two vector weights which 
were added to satisfy Equation~\ref{mayer_deconvolution}, and an updated 
expression for $\Phi_3$ developed by Tarazona~\cite{tarazonaphi3, santos2012phi3}, 
the excess free energy becomes
\begin{displaymath}{F_{excess}=k_{B}T\left(-n_{0}ln(1-n_{3})+\frac{n_{1}n_{2}-\vec{n_{1}}\cdot\vec{n_{2}}}{1-n_{3}}\color{white}\right)}\end{displaymath}
\begin{equation}{\color{white}\left(\color{black}+\frac{{n_2}^3-3n_2\vec{n}_{v2}\cdot\vec{n}_{v2}+\frac{9}{2}[\vec{n}_{v2}\cdot{\overleftrightarrow{n}_{m2}}\cdot{\vec{n}_{v2}}-\operatorname{Tr}({\overleftrightarrow{n}^3_{m2}})]}{24\pi(1-n_3)^2}\right)\color{black}V}\end{equation} 
where the tensor weight is given by
\begin{equation}\label{eq:tensorweight}{\overleftrightarrow{w}_{m2}(\vec{r})=w_2(r)\left(\frac{\vec{r}\vec{r}}{r^2}-\frac{I}{3}\right)}\end{equation} 
%and $\Phi_3$ is now
and
\begin{align}
\Phi_1 &= -n_{0}ln(1-n_{3}) \label{eq:Phi1}\\
\Phi_2 &= \frac{n_{1}n_{2}-\vec{n_{1}}\cdot\vec{n_{2}}}{1-n_{3}} \label{eq:Phi2} \\
\Phi_3 &= \frac{{n_2}^3-3n_2\vec{n}_{v2}\cdot\vec{n}_{v2}+\frac{9}{2}[\vec{n}_{v2}\cdot{\overleftrightarrow{n}_{m2}}\cdot{\vec{n}_{v2}}-\operatorname{Tr}({\overleftrightarrow{n}^3_{m2}})]}{24\pi(1-n_3)^2}  \label{eq:Phi3}
\end{align}

>>>>

%A number density profile is a time-invariant average 
%of the spacial distribution of the particles that make up the fluid, and
%There will be one, and only one 
%that minimizes the Helmholtz free energy.
%The minimum free energy, 
%is the value of the Helmholtz free energy that
%the system will have when it is in equilibrium at that temperature, 
%and its corresponding number density profile 
%are the equilibrium values the system will have at that temperature.
%is the equilibrium number density profile. 
%
Once an expression for the Helmholtz free energy as a functional of the
number density profile $n(\vec r)$ is formed then, according to classical 
Density Functional Theory, the equilibrium value of the 
Helmholtz energy, and its cooresonding equilibrium density profile can be
found by varying the number density profile
%can be varied 
through all possible configurations at some fixed temperature
until the free energy is minimized.  
In this section 
%To form the functional for the Helmholtz free energy, expressions for 
a functional for the Helmholtz free energy was formed by finding 
general expressions for the ideal and the excess portions of the free 
energy, and adding them together.
\begin{equation}{F[n(\vec{r})]=F_{ideal}[n(\vec{r})] + F_{excess}[n(\vec{r})]}\end{equation}
The functional for the ideal portion of the Helmholtz free energy is given by 
Equation~\ref{eq:Fideal-n(r)}, and a functional for the excess portion of the Helmholtz 
free energy is given in Equation~\ref{eq:Fexcess-simplified}. The specific 
form for the excess Helmholtz free energy depends on the particular 
fluid being modeled as defined by the  
interatomic potential energy V(r) between the spheres, which can make
forming an expression difficult. In this paper, Soft Fundamental Measure
Theory will be used to form an epxression for the excess Helmholtz free 
energy, as explained in the next section.

\section{Soft Fundamental Measure Theory (SFMT)}
In order to apply classical Density Functional Theory to a system that has
a fixed temperature, volume, and nubmer of particles, a functional for 
the Helmholtz free energy must first be formed. An expression for the ideal 
portion of the Helmnoltz free energy is given by Equation~\ref{eq:Fideal-n(r)}, 
but forming an expression for the excess Helmholtz free
energy for the particular fluid being modeled 
from Equation~\ref{eq:Fexcess-simplified} can be challenging. 
Soft Fundamental Measure Theory (SFMT) is a method for forming an 
expression for the excess Helmholtz free energy of a Soft-sphere fluid,
and will be adapted in the Methods section of this paper to fit the 
soft-sphere WCA fluid.
SFMT is an extension of Fundamental Measure Theory (FMT)
that was developed for the Hard-sphere fluid, and which can be derived from
Scaled Particle Theory (SPT).  
%For the Hard Sphere fluid, there is another way to form an expression for 
%the excess Helmholtz free energy in terms of a homogeneous number density 
%and, that is to use Scaled Particle Theory (SPT)~\cite{ReissSPT}.

Scaled Particle Theory seeks a general expression for the excess chemical 
potential of a homogeneous Hard Sphere fluid which can then be integrated 
to find the excess Helmholtz free energy according to the relation
%\begin{equation}F_{ex}=\int{\mu_{ex}dN}\end{equation}
%\begin{equation}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}\end{equation}
\begin{equation}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}{~~~~~}\rightarrow{~~~~~}F_{ex}=\int{\mu_{ex}dN}\end{equation}
It is an extension of the work done by Widom who showed that the excess 
chemical potential of a homogenesous fluid corredsponds to the reversible 
work of inserting a sphere of radius R into a fluid consisting of hard 
spheres of radius R. 
\begin{displaymath}\mu_{ex}=W_{rev}(R){~~~~}\text{work to insert a hard sphere of radius R}\end{displaymath}
Widom's result was obtained by considering the difference 
in the excess Helmholtz free energy when one atom is added to an otherwise 
closed system. He showed that the excess chemcial potential is related to 
the ensemble average of the Boltzman factor for the potential energy 
difference $V_{diff}$ between a system with N atoms and one with N+1 atoms. 
The resulting equation is known as Widom's Insertion Formula.
%\begin{displaymath}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}{~~~~~~}\end{displaymath}
\begin{equation}\mu_{ex}=\frac{F_{ex}(N+1)-F_{ex}(N)}{\Delta{N}}\bigg|_{T,V}\end{equation}
\begin{equation}\label{widoms-insertion-formula}{~}=-k_BT\ln\left(<e^{-\beta{V_{diff}}}>\right)\end{equation}
Considering that at each point in the fluid, there is a probability that 
either a hard sphere of radius R could be inserted, or not, and using 
these two probablities to evaluate $<e^{-\beta{V_{diff}}}>$ this becomes
\begin{equation}{~}=-k_BT\ln\left(e^{-\beta{V_{diff}}}+1\right)\end{equation}
\begin{equation}{~}=V_{diff}=W_{rev}(R)\end{equation}
Returning to Widom's insertion formula Equation~\ref{widoms-insertion-formula},
and considering that it takes no energy for a hard sphere to be placed at a point
in the fluid where it does not overlap with any other spheres, but would require
infinite energy to place a sphere where it would overlap another sphere, 
Equation~\ref{widoms-insertion-formula} becomes
\begin{equation}\mu_{excess}=-k_BT\ln(1-n\frac{4\pi}{3}R_{cavity}^3)\end{equation}
where the probability of a position existing in a homogeneous Hard-sphere fluid 
with number density n that a hard sphere of radius R can be placed without 
overlapping another sphere is given by the number density times the volume 
of the sphere, $nV_{Sphere}$.

%where the radius of the cavity $R_{cavity}$ is equal to the radius of a 
%hard sphere, R, and $<e^{-\beta{V_{diff}}}>$ is evaluated using the 
%probablities that the cavity is either filled or not filled~\cite{Hansen}.
%For cavities that are much larger than one of the hard spheres, 
Forming a more general result for different sizes of cavities in the fluid, 
Scaled Partical Theory arrives at an expression for the excess Helmholtz 
free energy of a homogeneous fluid~\cite{Hansen} in terms of the packing fraction $\eta$ 
\begin{equation}\label{Fexcess-SPT}{\frac{\beta{F_{excess}}}{N}=-ln(1-\eta)+\frac{3\eta}{1-\eta}+\frac{3{\eta}^2}{2(1-\eta)^2}}\end{equation} 
where
\begin{equation}{\eta = \frac{\mbox{Volume occupied by the spheres}}{\mbox{Total Volume}}}\end{equation}
\begin{equation}{\eta = \frac{N\frac{4}{3}\pi{R^3}}{V}}\end{equation}

@@

Incorporating into Equation~\ref{FexfromSPT} the two vector weights which 
were added to satisfy Equation~\ref{mayer_deconvolution}, and an updated 
expression for $\Phi_3$ developed by Tarazona~\cite{tarazonaphi3, santos2012phi3}, 
the excess free energy becomes
\begin{equation}{F_{excess} = k_BT(\Phi_1+\Phi_2+\Phi_3)V}\end{equation}
where
\begin{align}
\Phi_1 &= -n_{0}ln(1-n_{3}) \label{eq:Phi1}\\
\Phi_2 &= \frac{n_{1}n_{2}-\vec{n_{1}}\cdot\vec{n_{2}}}{1-n_{3}} \label{eq:Phi2} \\
\Phi_3 &= \frac{{n_2}^3-3n_2\vec{n}_{v2}\cdot\vec{n}_{v2}+\frac{9}{2}[\vec{n}_{v2}\cdot{\overleftrightarrow{n}_{m2}}\cdot{\vec{n}_{v2}}-\operatorname{Tr}({\overleftrightarrow{n}^3_{m2}})]}{24\pi(1-n_3)^2}  \label{eq:Phi3}
\end{align}
and the tensor weight is given by
\begin{equation}\label{eq:tensorweight}{\overleftrightarrow{w}_{m2}(\vec{r})=w_2(r)\left(\frac{\vec{r}\vec{r}}{r^2}-\frac{I}{3}\right)}\end{equation}
